Demes: a specification for interchanging complex demography

### 2021-06-24 paper.tex

\documentclass[11pt]{article}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{booktabs}

@@

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly increasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few (ideally, one
per programming language), high-quality Demes parsers available as
libraries the probability of divergences from the standard is
greatly reduced.

The Demes specification precisely defines the required behaviour of
parsers, and we also provide a reference implementation written
in Python to resolve any potential ambiguities and to provide a
helpful template for other implementations. We also provide an
extensive suite suite of examples and the expected outputs. In addition,
we have high-quality parser implementations in the Python and C
languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

The Demes specification is by design simple and static---we wish to
unambiguously describe a demographic model with fully specified
and fixed parameters. This simplicity means that we cannot directly
specify distributions or estimated confidence intervals
for those parameters. While it is not difficult to imagine extending
the specification in ways that would allows us to do this, it would
greatly increase the complexity of parsers and would likely be of
little practical use in any case. This is because the parameters of
demographic models are typically tightly coupled, and cases in which
distributions for different parameters can be simply described are rare.
In this situation, the simplest way to describe estimated
distribution is simply to list a large number of samples from
the posterior. While writing out a large number of Demes models in
YAML format may seem inefficient, it can in fact be a compact
way to describe these distributions.
\jkcomment{TODO: add in a quick example here where we take an IM model
and simulate a posterior distribution under some reasonable model. Report
how much space it takes to represent a large number of samples from this
distribution}

>>>>

\documentclass[11pt]{article}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}

@@

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few (ideally, one
per programming language), high-quality Demes parsers available as
libraries the probability of divergences from the standard is
greatly reduced.

The Demes specification precisely defines the required behaviour of
parsers, and we also provide a reference implementation written
in Python to resolve any potential ambiguities and to provide a
helpful template for other implementations. We also provide an
extensive test suite of examples and the expected outputs. In addition,
we have high-quality parser implementations in the Python and C
languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

The Demes specification is by design simple and static---we wish to
unambiguously describe a demographic model with fully specified
and fixed parameters. This simplicity means that we cannot directly
specify distributions or estimated confidence intervals
for those parameters. While it is not difficult to imagine extending
the specification in ways that would allows us to do this, it would
greatly increase the complexity of parsers and would likely be of
little practical use in any case. This is because the parameters of
demographic models are typically tightly coupled, and cases in which
distributions for different parameters can be simply described are rare.
In this situation, the simplest way to describe estimated
distribution is simply to list a large number of samples from
the posterior. While writing out a large number of Demes models in
YAML format may seem inefficient, it can in fact be a compact
way to describe these distributions.
For example, consider a one-population model with piecewise-constant sizes over
20 epochs which has $\sim$~40 free parameters---the \texttt{start\_size} and
\texttt{end\_time} values for each epoch. Supposing we sample 50,000 models
from the posterior distribution, the resulting multi-document YAML file is
45~MiB.
This format compresses down to 8.4~MiB when gzipped or 6.2~MiB
when compressed with LZMA2, which is on par with an equivalent binary
representation of the free parameters
($40\times50000\times4~\text{bytes} \approx 7.6~\text{MiB}$).
